Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-24 paper/paper.tex
Clean up paper paper formatting.

\documentclass[10pt]{article}

@@

%\setlength{\parskip}{\baselineskip}     % skip line following paragraphs
\setlength{\parskip}{0.0in}
\setlength{\parindent}{0.3in}           % Control margins and amount of text
\setlength{\topmargin}{-0.8in}
\setlength{\oddsidemargin}{-.15in}      % changed from {-.15in}
\setlength{\textheight}{9.5in}
\setlength{\textwidth}{6.8in}
\pagestyle{empty}                       % No page numbers

@@

{\LARGE\bf
\begin{tabbing}
\hspace{2.8in} \= \hspace{1.3in} \= \hspace{1.2in} \= \\

%=========    Heading   ==================================================
CS 399 \> Paper \> Bill Chickering (bchick)\\
\normalsize Mar 21, 2014 \> \> Jamie Irvine (jirvine)
% =========================================================================
\end{tabbing}
} \vspace{.4in}

@@

There are a few ways to compute the similarity of two vectors. One approach is
to use Cosine-similarity, defined as the cosine of the angle between the two vectors:
\begin{align}
CosSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{A}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{B}} r_{u,B}^2}}
\end{align}
where $U_{I}$ is the set of all users who rated item $I$, $U_{AB}$ is the set of all 
users who rated both item $A$ and item $B$ and $r_{u,I}$ is the rating user $u$ gave 
to item $I$. Another popular approach is to use Pearson correlation:
\begin{align}
PearsSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{AB}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{AB}} r_{u,B}^2}}
\end{align}
Other similarity functions exist, such as one-sided similarities, but Pearson
correlation and Cosine-similarity are the most popular and will be the two
measurements used in this paper. \footnotemark

@@

\section*{Problem}
The goal is to calculate a modified similiary function that leverages the number of
commonn users to better estimate the true similarity. Formally, we design
$ModSim$ such that for two items $A$ and $B$ with $n$ users in common
\begin{align}
ModSim(A, B, n) \approx TrueSim(A, B)
\end{align}
where $TrueSim$ is the true similarity of the two items. Of course, there is no
way to actually know the true similarity of two items, but for a good enough
similarity function, such as $PearsSim$ or $CosSim$ and a large enough $n$, we
can get a good estimate. Thus, we model $TrueSim$ as 
\begin{align}
TrueSim(A, B) = lim_{n\to\infty}Sim(A, B)
\end{align}
where $Sim$ is the most appropriate generic similarity function for the dataset.

Since $TrueSim$ dependends on the choice of $Sim$, we abstract away the details
of $Sim$ and use its output directly. In all, the goal is to construct $ModSim$ such that
\begin{align}
ModSim(Sim(A, B), n) \approx lim_{n\to\infty}Sim(A, B)
\end{align}

\section*{Model}
We model the problem probabilistically. Let $Y$ be a random variable
representing the $TrueSim$ of a randomly chosen pair of items. Let $X_n$ be a
random variable representing the calculated $Sim$ of a pair of items with $n$ common users.

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{1}^2)
\end{align}
where $\mu$ and $\sigma_{1}^2$ are the average similarity score and the variance of similarity scores
of all pairs of items, respectively. Since $X_n$ represents a noisy reading of
the true similarity $Y$, we model $X_n | Y$ as a Normal distribution around $Y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{2, n}^2)
\end{align}
Note that $\sigma_{2, n}^2$ represents how noisy the estimation of $Sim$ is
when there are $n$ common users. This should decrease as $n$ increases.

@@

In probabilistic terms, for a calculated value $X_n$, we want to find the value 
of $Y$ that most likely produced $X_n$. Thus we desire the Maximum Likelihood
Estimate of $Y | X_n$
\begin{align}
\argmax_yP(Y=y|X_n=x) &= \argmax_y\left[\frac{P(X_n=x |
Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
Solving for y:
\begin{align}
y &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(y - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}\left(x-\mu\right)
\end{align}

@@

The parameters $\mu$, $\sigma_{1}^2$, and $\sigma_{2,n}^2$ all can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{2,n}^2$ for every value of $n$ separately. Alternatively, we can model
$\sigma_{2,n}^2$ as a function of $n$. Recall that $\sigma_{2,n}^2$ is the
noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{2,n}^2$ as:

\begin{align}
\sigma_{2,n}^2 = \frac{\alpha}{\sqrt{n}}
\end{align}
Yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}

>>>>

\documentclass[11pt]{article}

@@

%\setlength{\parskip}{\baselineskip}     % skip line following paragraphs
%\pagestyle{empty}                       % No page numbers
\setlength{\topmargin}{-.5in}
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{.125in}
\setlength{\textwidth}{6.25in}

@@

\title{Improving Item-Item Similarity Estimation during Collaborative Filtering}
\author{Bill Chickering and Jamie Irvine\\
CS 399 with Anand Rajaraman\\
Stanford University}
\renewcommand{\today}{March 25, 2014}
\maketitle

@@

There are a few ways to compute the similarity of two vectors. One approach is
to use Cosine-similarity, defined as the cosine of the angle between the two
vectors:
\begin{align}
CosSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{A}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{B}} r_{u,B}^2}}
\end{align}
where $U_{I}$ is the set of all users who rated item $I$, $U_{AB}$ is the set of
all users who rated both item $A$ and item $B$ and $r_{u,I}$ is the rating user
$u$ gave to item $I$. Another popular approach is to use Pearson correlation:
\begin{align}
PearsSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{AB}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{AB}} r_{u,B}^2}}
\end{align}

Other similarity functions exist, such as one-sided similarities, but Pearson
correlation and Cosine-similarity are the most popular and will be the two
measurements used in this paper. \footnotemark

@@

\section*{Problem}

The goal is to calculate a modified similiary function that leverages the number
of commonn users to better estimate the true similarity. Formally, we design
$ModSim$ such that for two items $A$ and $B$ with $n$ users in common

\begin{align}
ModSim(A, B, n) \approx TrueSim(A, B)
\end{align}
where $TrueSim$ is the true similarity of the two items. Of course, there is no
way to actually know the true similarity of two items, but for a good enough
similarity function, such as $PearsSim$ or $CosSim$ and a large enough $n$, we
can get a good estimate. Thus, we model $TrueSim$ as 
\begin{align}
TrueSim(A, B) = lim_{n\to\infty}Sim(A, B)
\end{align}
where $Sim$ is the most appropriate generic similarity function for the dataset.

Since $TrueSim$ dependends on the choice of $Sim$, we abstract away the details
of $Sim$ and use its output directly. In all, the goal is to construct $ModSim$
such that

\begin{align}
ModSim(Sim(A, B), n) \approx lim_{n\to\infty}Sim(A, B)
\end{align}

\section*{Model}

We model the problem probabilistically. Let $Y$ be a random variable
representing the $TrueSim$ of a randomly chosen pair of items. Let $X_n$ be a
random variable representing the calculated $Sim$ of a pair of items with $n$
common users.

We model $Y$ as a Normal distribution 
\begin{align}
Y \sim N(\mu, \sigma_{1}^2)
\end{align}
where $\mu$ and $\sigma_{1}^2$ are the average similarity score and the variance
of similarity scores of all pairs of items, respectively. Since $X_n$ represents
a noisy reading of the true similarity $Y$, we model $X_n | Y$ as a Normal
distribution around $Y$:
\begin{align}
(X_n | Y=y) \sim N(y, \sigma_{2, n}^2)
\end{align}

Note that $\sigma_{2, n}^2$ represents how noisy the estimation of $Sim$ is when
there are $n$ common users. This should decrease as $n$ increases.

@@

In probabilistic terms, for a calculated value $X_n$, we want to find the value
of $Y$ that most likely produced $X_n$. Thus we desire the Maximum Likelihood
Estimate of $Y | X_n$
\begin{align}
\argmax_yP(Y=y|X_n=x) &= \argmax_y\left[\frac{P(X_n=x |
Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
Solving for y:
\begin{align}
y &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(y - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

@@

The parameters $\mu$, $\sigma_{1}^2$, and $\sigma_{2,n}^2$ all can be
approximated using the method of moments over training data where there are
enough common users that $TrueSim$ can be approximated. This method requires
approximating $\sigma_{2,n}^2$ for every value of $n$ separately. Alternatively,
we can model $\sigma_{2,n}^2$ as a function of $n$. Recall that $\sigma_{2,n}^2$
is the noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n
\to \infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{2,n}^2$ as:
\begin{align}
\sigma_{2,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}
